Parameter Space Representation of Pareto Front to Explore Hardware-Software Dependencies

### 2014-10-29 algorithm.tex
Added table on design space

\subsection{Exploration vs exploitation in PS}
Our algorithm is able to inherently balance exploration and exploitation, which are ``two cornerstones of problem solving by search''~\cite{eiben1998evolutionary}. While a detailed survey on the exploration vs exploitation trade-off can be found in~\cite{vcrepinvsek2013exploration}, here we limitedly recall that \emph{exploration} is the capability of an algorithm to consider different parts of the configuration space, not being confined only to a limited subset and being trapped in local optima. Following the terminology of \cite{weise2012evolutionary}, if the exploration is not sufficiently wide, the algorithm suffers from \emph{premature convergence}, i.e. it starts to analyze only a narrow portion of the configuration space, thus ignoring potential Pareto configurations that are outside this portion. When premature convergence happens, there is insufficient \emph{diversity}~\cite{weise2012evolutionary}, i.e. all the provided solutions are similar.
On the other hand, \emph{exploitation} is the capability to recognize and densely analyze the parts of configuration space that have many Pareto configurations.

@@

To summarize,thanks to the above characteristics, PS is able to recognize the regions when the best configurations are, while avoiding premature convergence, not being trapped in a narrow subset of configuration space.

@@

\comment{TODO: Usare notazioni coerenti con le precedenti}
As regards the performance of the proposed approach we can estimate that the
performance of $PS$ is strictly related to the total number of
different configurations evaluated during the exploration. We can
safely make such assumption, because the process of evalutating a new
configuration is the time-dominating one among all the operations
performed. As we will show the
next Section, this is especially true in the hardware/software
design space exploration scenario we are addressing, since benchmark compiling and execution times
are orders of magnitude bigger than the splitting/merging operations and
innovation scores computations.  Ideally, the $i$-th era would provide a new set
$test_i$ of $K$ configurations to be tested. However, we cannot know
a priori how many of
these will actually be new configurations, i.e. requiring a process of
evaluation. Thus, number of total new simulations $N_{tot}$ 
can be estimated to be bounded as follows:
\[
K \le N_{tot} \le K \times E_{tot}
\]
where $E_{tot}$ is the nuber of eras executed and assuming that at
least $K$ evaluation should be performed in order to obtain the first
pareto $\mathscr{P}_0$.

\subsection{On the parallelizability of PS}
While the detailed analysis and the evaluation of a parallel implementation of our algorithm is not in the scope of this paper, we provide in this section some arguments to show that PS is parallelizable.

@@

Another possible approach is closer to the ``Island scheme'' for GA parallelization~\cite{Borkar14parallel}. The configuration space is initially divided in $n$ macro-regions, each one assigned to a different processing node, which autonomously performs all the PS steps, including merging and splitting, considering the assigned macro-region as the entire parameter space. Note that, while the Island scheme for GA parallelization requires a migra0tion operator, that is not easy to properly tune, we do not have special requirements. On the other hand, one of the macro-regions may begin to provide very low innovation at a certain era. In this case, dedicating to it the same number $K$ of test function as the other macro-regions is inefficient. To solve this problem, the number of test functions assigned to each macro-region should be perdiodically renegotiated, in favor of the macro-regions providing most innovation.

>>>>

\subsection{Exploration vs Exploitation in PS}
The proposed algorithm has been designed in order to be inherently able to balance exploration and exploitation, which are ``two cornerstones of problem solving by search''~\cite{eiben1998evolutionary}. While a detailed survey on the exploration vs exploitation trade-off can be found in~\cite{vcrepinvsek2013exploration}, here we limitedly recall that \emph{exploration} is the capability of an algorithm to consider different parts of the configuration space, not being confined only to a limited subset and being trapped in local optima. Following the terminology of \cite{weise2012evolutionary}, if the exploration is not sufficiently wide, the algorithm suffers from \emph{premature convergence}, i.e. it starts to analyze only a narrow portion of the configuration space, thus ignoring potential Pareto configurations that are outside this portion. When premature convergence happens, there is insufficient \emph{diversity}~\cite{weise2012evolutionary}, i.e. all the provided solutions are similar.
On the other hand, \emph{exploitation} is the capability to recognize and densely analyze the parts of configuration space that have many Pareto configurations.

@@

\comment{CHECK AND: Per caso era where invece di when ? o manca qualcosa ?}
To summarize,thanks to the above characteristics, PS is able to recognize the regions when the best configurations are, while avoiding premature convergence, not being trapped in a narrow subset of configuration space.

@@

As regards the performance of the proposed approach we can estimate that the
performance of $PS$ is strictly related to the total number of
different configurations evaluated during the exploration. We can
safely make such assumption, because the process of evalutating a new
configuration is the time-dominating one among all the operations
performed. As we will show the
next Section, this is especially true in the hardware/software
design space exploration scenario we are addressing, since benchmark compiling and execution times
are orders of magnitude bigger than the splitting/merging operations and
innovation scores computations.  Ideally, the $i$-th era would provide a new set
$test_i$ of $K$ configurations to be tested. However, we cannot know
a priori how many of
these will actually be new configurations, i.e. requiring a process of
evaluation. Thus, number of total new simulations $N_{tot}$ 
can be estimated to be bounded as follows:
\[
K \le N_{tot} \le K \times E_{tot}
\]
where $E_{tot}$ is the nuber of eras executed and assuming that at
least $K$ evaluation should be performed in order to obtain the first
pareto $\mathscr{P}_0$.

\subsection{On the parallelizability of PS}
While the detailed analysis and the evaluation of a parallel
implementation of the proposed algorithm is not in the scope of this paper, we
provide in this subsection some arguments to show different strategies
that could be applied to make PS execution parallelizable.

@@

Another possible approach is closer to the ``Island scheme'' for GA
parallelization~\cite{Borkar14parallel}. The configuration space is
initially divided in $n$ macro-regions, each one assigned to a
different processing node, which autonomously performs all the PS
steps, including merging and splitting, considering the assigned
macro-region as the entire parameter space. Note that, while the
Island scheme for GA parallelization requires a migration operator,
that is not easy to properly tune, we do not have special
requirements. On the other hand, one of the macro-regions may begin to
provide very low innovation at a certain era. In this case, dedicating
to it the same number $K$ of test function as the other macro-regions
is inefficient. To solve this problem, the number of test functions
assigned to each macro-region should be perdiodically renegotiated, in
favor of the macro-regions providing most innovation.

### 2014-10-29 experiments.tex
related works fixes
fixed deps descr
Added table on design space

\section{A Case Study: VLIW Hardware/Software Design Space}
\secL{ee}
Although PS has been formally described as a general exploration algorithm, the goal of this work is giving a method applicable to the embedded system design. For this reason, we evaluate PS in this context. To this aim, we use a parameterized Very Long Instruction
Word (VLIW) architecture~\cite{kathail_tr00} as testbed. The philosophy
behind a VLIW system makes this a natural choice for several reasons:
\begin{itemize}
\item \emph{Multi-Objective trade-offs}: VLIW architectures allow designers
to trade between power, energy and performances objectives, resulting
in extended Pareto sets which are an ideal ground for our testing
purposes focused on parameter-space representation
\item \emph{Compiler awareness}: code scheduling for the execution of
applications is statically obtained by the compiler which
performs code transformations ``being aware'' of the underlying hardware
configuration.  This tight hardware/software dependence makes the
VLIW scenario perfectly suitable for investigating hardly predictable
parameter interactions.
\end{itemize}

@@

\subsection{Design Space: Hardware and Software Parameters}
Conceptually, system parameters can be classified in three main categories:
\emph{processor}, \emph{memory sub-system} and \emph{compiler}. The
first two categories are directly related to the physical
implementation of the system, so parameters of those
classes are usually referred ``hardware parameters''. On the other
side, all the parameters affecting the behaviour of the compiler
are referred as ``software parameters'', being directly involved in
the process of generating the application executables running on the
underlying hardware.

Processor hardware parameters are directly related to the VLIW core and includes the size
of the register files and the number of functional
units of each type. In particular, five different types of register
files are supported:
GPR (32-bit registers for integers), FPR (64-bit registers for
floating point values) PR (1-bit registers used to store the Boolean
values of predicated instructions), BTR (64-bit registers containing
information about possible future branches) and CR (32-bit control
registers containing information about the internal state of the
processor). The functional units involved are: \emph{Integer units},
\emph{floating point units}, \emph{memory units} (associated with
load/store operations) and \emph{branch units} (associated with branch
operations). With respect to the memory sub-system hardware, the parameters
include \emph{size}, \emph{associativity} and
\emph{block size} for each of the three caches: First-level data cache
(L1D), first-level instruction cache (L1I) and second-level unified
cache (L2U).

On the software side, a set of the most impacting code
compilation parameters has been selected and included in the
configuration space, in particular:
\begin{itemize}
\item \emph{tcc\_region}: Specifies the scope of action of the compiler and the type
of code transformation involved (basic block, super block and hyper
block) 
\item \emph {max\_unroll\_allowed}: The number of unroll iterations allowed
\item \emph{regroup\_only}: Avoids inlining 
\item \emph{do\_classic\_opti}: A set of classical optimization, not VLIW related,
such as common expression removal and loop unrolling
\item \emph{do\_prepass\_scalar\_scheduling}: Performs a schedule before
forming regions 
\item \emph{do\_postpass\_scalar\_scheduling}: Performs a schedule after the region formation 
\item \emph{do\_modulo\_scheduling}: Modular scheduling of loop code
\item \emph{memvr\_profiled}: Performs a memory-dependencies profiling 
\end{itemize}

@@

% ACM table style
\begin{table}
	\tbl{Benchmarks\label{tab:bench}}{
	\begin{tabular}{ll}
	\hline
	\multicolumn{1}{c}{Benchmark} & \multicolumn{1}{c}{Application} \\
	\hline
	Alloca\_test & Memory array allocation test \\
	Bmm & Matrices multiplication and elements sum \\
	Fir\_int & Finite impulse response \\
	Mm & Floating point matrices multiplication \\
	Sha256 & Cryptocurrency header Hashing \\
	Wave & Wavefront computation \\
	\hline
	\end{tabular}}
\end{table}

%------------------------------------------------------------------------------
\subsection{Configuration Evaluation Flow}

@@

\comment{TODO controllare inglese}
Even if an analogy between the concepts of ``era'' and ``generation'' 
can be considered, a fixed amount of total different simulations has been
considered for each of the tests. The purpose is to make the
comparison fair from a algorithm performance perspective, since a
fixed budget of simulations is directly related to the amount of time
required to execute the exploration; on the other side, just simply
referring to a given number of generations/era could be misleading when a
significative number of configuration remain unaltered in the current
population (e.g. because of a local optimum).

The set of default algorithm tuning parameters used for both the
genetic and PS approaches to perform the simulations are shown in
Table~\ref{tab:defaults}. Given the similarity, both in terms of
meaning and functionality, of the parameter $K$ with the population size
of the genetic approach, they both have been set to the same values. As discussed
when introducing the main PS algorithm concepts, an accurate
investigation of algorithm tuning is beyond the scope of this
introductory work. Nevertheless, a set of preliminary tests with
different values of $K$ have been conducted. As shown in
Figure~\ref{fig:kimpact}, which refer to the alloca\_test application
and a fixed amount of 500 configurations, initially increasing values
of $K$ seem to improve both pareto (left picture) and objective
distribution (right picture), but after some threshold the $K$ does
not seem to provide any benefit, returning to slightly worse
results for the last value considered (100). Similar trends have been
found for other benchmarks in the set considered (which are
not included for space reasons), showing a value of $K$ around 30 as
reasonable sweet spot for the objectives. A complete resource with the whole 
results and data obtained in the tests performed can be found
at~\cite{ps_results}.

@@

\subsection{Evaluation metrics}
In order to evaluate the Paretos from a quantitative point-of-view we
will consider two metrics, namely, the \emph{variation range} and the
\emph{average normalised absolute dispersion error}. Both of them permit to quantify the extent of the Pareto front that is an important factor to consider when evaluating the quality of a Pareto front. Indeed, authors in \cite{zitzler_ec00,weise2012evolutionary} observe that an exploration algorithm should preferably provide Pareto fronts whose objective function range is large. In other words, following the terminology of~\cite{weise2012evolutionary}, a \emph{uniform convergence} should be guaranteed.
Pareto fronts with large objective function ranges permit to investigate in a broader way the potentials of the designed device. This is important, in general, since if only a narrow portion of the Pareto front is provided due to non-uniform convergence, the decision maker may neglect some important options~\cite{weise2012evolutionary}. Having a wide Pareto front becomes essential, in particular, when designing a device that may be required to satisfy different constraints. For example, the decision maker may want to design two different multi-purpose processors. The first design may adapt to situations in which the processor is used for real time elaboration (thus requiring low execution time) and is attached to the powerline (thus not having stringent power dissipation constraints). The second design, on the contrary, may fit situations in which the energy is provided by a battery (stringent power constraints) but only background computations are performed (long execution time is tolerable). These two processors may be successfully designed maintaining the same architecture, using different values for the design parameters. A wide Pareto front helps in achieving this goal.

@@

\comments{Da lasciare, da spostare ? da mostrare in modo semplice ?
non so}
Another interesting feature of PS is that the final output of its
execution is not only the set of pareto optimal configurations, but also a set
of more or less innovation regions of the parameter space, obtained as
a result of the merge/splitting operators described
in~\secR{algorithm}.

>>>>

\section{A Case Study: VLIW Hardware/Software Design Space}
\secL{ee}
Although PS has been formally introduced and described as a general
multi-objective optimization algorithm, in this work we will
explicitly refer to the more restricted (but still very large)
embedded system design scenario,  since its properties
are the one that suggested to the authors the concepts of pareto-based
innovation and region scoring system.

In particular, we will use a parameterized Very Long Instruction
Word (VLIW) architecture~\cite{kathail_tr00} as testbed. The philosophy
behind a VLIW system makes this a natural choice for several reasons:
\begin{itemize}
\item \emph{Multi-Objective trade-offs}: VLIW architectures allow designers
to trade between power, energy and performances objectives, resulting
in extended Pareto sets which are an ideal ground for our testing
purposes focused on parameter-space representation
\item \emph{Software Compiler awareness}: code scheduling for the execution of
applications is statically obtained by the compiler which
performs code transformations ``being aware'' of the underlying hardware
configuration.  This tight hardware/software dependence makes the
VLIW scenario perfectly suitable for investigating hardly predictable
parameter interactions.
\end{itemize}

@@

\subsection{Design Space: Hardware and Software Parameters}
\secL{designspace}
Conceptually, system parameters can be classified in three main categories:
\emph{processor}, \emph{memory sub-system} and \emph{compiler}. The
first two categories are directly related to the physical
implementation of the system, so parameters of those
classes are usually referred as ``hardware parameters''. On the other
side, all the parameters affecting the behaviour of the compiler
are referred as ``software parameters'', being directly involved in
the process of generating the application executable running on the
underlying hardware.

Processor hardware parameters are directly related to the VLIW core
and includes the size of the register files and the number of
functional units of each type. On the software side, a set of the most
impacting code compilation parameters has been selected.
Table~\ref{tab:param} show the complete set of parameters included in
the design space and their possible values. The total size of design
space, which grows with the product of parameter cardinalities is
about $1.33 \times 10^{14}$.

@@

\begin{table}
	\tbl{Design Space\label{tab:param}}{
	\begin{tabular}{lll}
	\hline
	\multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Values} & \multicolumn{1}{c}{Type}\\
	\hline
	General Purpose Registers & 16, 32, 64, 128 & HW \\
	Floating Point Registers & 16, 32, 64, 128 & HW \\
	Predicate Instructions registers & 32, 64, 128 & HW \\
	Branch Target Registers & 16, 32, 64 & HW \\
	Control Registers & 32, 64, 128 & HW \\
	\hline
	Integer Functional Units & 1, 2, 3, 4, 5, 6 & HW \\
	Floating Point Units & 1, 2, 3, 4, 5, 6 & HW\\
	Memory Units & 1, 2, 3, 4  & HW \\
	Branch Units & 1, 2, 3, 4  & HW \\
	\hline
	L1 Data Cache Size (KB) & 1, 2, 4, 8, 16, 32, 64, 128 & HW \\
	L1 Data Cache Blocksize & 32, 68, 128 & HW \\
	L1 Data Cache Associativity (KB) & 1, 2, 4 & HW \\
	L1 Instruction Cache Size (KB) & 1, 2, 4, 8, 16, 32, 64, 128 & HW \\
	L1 Instruction Cache Blocksize & 32, 68, 128 & HW \\
	L1 Instruction Cache Associativity (KB) & 1, 2, 4 & HW \\
	L2 Unified Cache Size (KB) & 32, 64, 128, 256, 512  & HW\\
	L2 Unified Cache Blocksize & 32, 68, 128 & HW \\
	L2 Unified Cache Associativity (KB) & 2, 4, 8 & HW \\
	\hline
        tcc\_region: Scope of action of the compiler & basic block, super block, hyper block & SW \\
        max\_unroll\_allowed: unroll iterations allowed & none, medium, high & SW \\
        regroup\_only: Avoids inlining & yes, no & SW \\
        do\_classic\_opti: Classical optimizations & yes, no & SW \\
        do\_prepass\_scalar\_scheduling: Performs a prepass schedule & yes, no & SW \\
        do\_postpass\_scalar\_scheduling: Performs a schedule after & yes, no & SW \\
	do\_modulo\_scheduling: Modular scheduling of loop code & yes, no & SW \\
        memvr\_profiled: Memory-dependencies profiling & yes, no & SW \\
	\hline
	\end{tabular}}
\end{table}

% ACM table style
\begin{table}
	\tbl{Benchmarks\label{tab:bench}}{
	\begin{tabular}{ll}
	\hline
	\multicolumn{1}{c}{Benchmark} & \multicolumn{1}{c}{Application} \\
	\hline
	Alloca\_test & Memory array allocation test \\
	Bmm & Matrices multiplication and elements sum \\
	Fir\_int & Finite impulse response \\
	Mm & Floating point matrices multiplication \\
	Sha256 & Cryptocurrency header Hashing \\
	Wave & Wavefront computation \\
	\hline
	\end{tabular}}
\end{table}
%------------------------------------------------------------------------------
\subsection{Configuration Evaluation Flow}

@@

Even if an analogy between the concepts of ``era'' and ``generation'' 
can be considered, a fixed amount of total different simulations has been
considered for each of the tests. The purpose is to make the
comparison fair from a algorithm performance perspective, since a
fixed budget of simulations is directly related to the amount of time
required to execute the exploration; on the other side, just simply
referring to a given number of generations/era could be misleading when a
significative number of configuration remain unaltered in the current
population (e.g. because of a local optimum).

The set of default algorithm tuning parameters used for both the
genetic and PS approaches to perform the simulations are shown in
Table~\ref{tab:defaults}. Given the similarity, both in terms of
meaning and functionality, of the parameter $K$ with the population size
of the genetic approach, they both have been set to the same values. As discussed
when introducing the main PS algorithm concepts, an accurate
investigation of algorithm tuning is beyond the scope of this
introductory work. Nevertheless, a set of preliminary tests with
different values of $K$ have been conducted, similar the one shown in
Figure~\ref{fig:kimpact} which refer to the alloca\_test application
and a fixed amount of 500 configurations. We found that initially increasing values
of $K$ seem to improve both pareto (as in the picture on the left) and objective
distribution (right picture), but after some threshold value the $K$ does
not seem to provide any benefit, even returning to slightly worse
results for the last value of $K$ considered (100). Similar trends have been
found for the other benchmarks (not included for space reasons),
showing on average a value of $K$ around 30 as
reasonable sweet spot for the objectives. A complete online resource with the whole 
results and data obtained in the tests performed can be found
at~\cite{ps_results}.

@@

\subsection{Evaluation metrics}
Qualitative evaluation of paretos is probably the simplest way for
having a quick glance on algorithm behviour and results. However, in
tests involving a comparison between different exploration strategies
that approach could not help in capturing the whole picture, e.g. missing
some important properties related to the distribution of the solutions
found.
In order to evaluate the Paretos from a quantitative point-of-view we
will consider two metrics, namely, the \emph{variation range} and the
\emph{average normalised absolute dispersion error}. Both of them permit to quantify the extent of the Pareto front that is an important factor to consider when evaluating the quality of a Pareto front. Indeed, authors in \cite{zitzler_ec00,weise2012evolutionary} observe that an exploration algorithm should preferably provide Pareto fronts whose objective function range is large. In other words, following the terminology of~\cite{weise2012evolutionary}, a \emph{uniform convergence} should be guaranteed.
Pareto fronts with large objective function ranges permit to investigate in a broader way the potentials of the designed device. This is important, in general, since if only a narrow portion of the Pareto front is provided due to non-uniform convergence, the decision maker may neglect some important options~\cite{weise2012evolutionary}. Having a wide Pareto front becomes essential, in particular, when designing a device that may be required to satisfy different constraints. For example, the decision maker may want to design two different multi-purpose processors. The first design may adapt to situations in which the processor is used for real time elaboration (thus requiring low execution time) and is attached to the powerline (thus not having stringent power dissipation constraints). The second design, on the contrary, may fit situations in which the energy is provided by a battery (stringent power constraints) but only background computations are performed (long execution time is tolerable). These two processors may be successfully designed maintaining the same architecture, using different values for the design parameters. A wide Pareto front helps in achieving this goal.

@@

\comment{Da lasciare, da spostare ? da mostrare in modo semplice ?  non so}
Another interesting feature of PS is that the final output of its
execution is not only the set of pareto optimal configurations, but also a set
of more or less innovation regions of the parameter space, obtained as
a result of the merge/splitting operators described
in~\secR{algorithm}.
